---
title: "Practical Machine Learning : Human Activity Recognition "
author: "Jacob Paracka"
date: "January 8, 2018"
output: html_document
---

# Introduction
The exercise aims to develop a prediction model which can recognize human activity of weight lifting and determine if some of the common mistakes are being made when performing the exercise.

```{r loadlibraries}
knitr::opts_chunk$set(echo = TRUE)
##load required libraries 
library(caret)
library(knitr)
library(rpart)
```

# Download and prepare data
```{r downloaddata}
training <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", na.strings = c("NA",""))

validation <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", na.strings = c("NA",""))

```

# Exploratory data analysis 

```{r explpratoryanalysis}
dim(training)
##head(training)
nrow(training[training$new_window=='yes',])
```

On observing the data we can already eliminate some of the fields that are not relevant for the prediction. Columns user_name, Timestamps and num_window are not relevant to the prediction. Also we for column new_window only 406 out of the 19622 records ae yes. There are also many columns where the values are mostly null. These columns can be removed for the dataset. 

#  Data preparation 
```{r datapreparation}
## remove the first 7 columns that are not relevant to prediction
training <- training[, -c(1:7)]
##remove fields with majority NA values 
df <- sapply(training, function(x) sum(is.na(x)))
training <- training[, (df<19000)]
dim(training)

##remove highy correrelated coolumns, columns that a corelation more than 0.95 are removed 
tmp <- cor(training[,1:52])
tmp[upper.tri(tmp)] <- 0
diag(tmp) <- 0
training.new <- training[, !apply(tmp,2,function(x) any(x > 0.95))]
training.new$classe <- training[,'classe']

dim(training.new)
```

The final dataset used in creating the model has 51 predictors.

#Split into training and testing data set 
```{r splitdata}
inTrain = createDataPartition(y=training.new$classe, p = 3/4)
train = training.new[inTrain[[1]],]
test = training.new[-inTrain[[1]],]
```

# Creating the prediction models 
This is a classification prediction. We choose the random forest model with  5 fold cross validation to build the predictor.

```{r predictionmodel}
## create 2 vectors x the predictors and y is the result
x <- train[,-52]
y <- train[, 52]


fitControl <- trainControl(method = "cv",number = 5, allowParallel = TRUE)
##fitrf <- train(x, y, method = "rf", data = training.new, trControl = fitControl)


## Test the accuracy of the model with the test data
predrf <- predict(fitrf, test)

## print the confustion matrix 
matrix2 <- confusionMatrix(predrf, test$classe)
matrix2$table
matrix2$overall

```

# Prediction accuracy 
The Model has an accuracy of 99.3% for the test data. The out of sample error is less than 0.6%. This is a good accuracy for prediction.

# Predicting from  test data 
Use the developed model to predict the outcomes for the testing dataset 

```{r}
pred <- predict(fitrf, validation)
data.frame(validation$problem_id, pred)

```

##end



